---
title: "Class 14 - Pathway Analysis"
author: "Peter Sax"
format: html
toc: true
---

## 1. Differential Expression Analysis

```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
```


> Q1. Remove the troublesome first column from `countData`.

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```
> Q2. Remove the 0 entries from the matrix

```{r}
countData = countData[rowSums(countData) > 0, ]
head(countData)
```

### Running DESeq2

We can now run the DESeq2 pipeline similar to last class...

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```

```{r}
dds
```
```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```

> Q3. Call the summary function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff

```{r}
summary(res)
```
### Volcano Plot

Now we will make a volcano plot, a commonly produced visualization from this type of data that we introduced last day. Basically it's a plot of log2 fold change vs -log adjusted p-value.

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

> Q4. Improve the plot by adding color and axis labels

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

### Adding Gene Annotation

Since we mapped and counted against the Ensembl annotation, our results only have information about Ensembl gene IDs. However, our pathway analysis downstream will use KEGG pathways, and genes in KEGG pathways are annotated with Entrez gene IDs. So lets add them as we did the last day.

> Q5. Use the `mapIDS()` function multiple times to add SYMBOL, ENTREZID, and GENENAME annotation to our results

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```
> Q6. Reorder the results by pvalue and save to a CSV file in the project directory

```{r}
res = res[order(res$pvalue),]
write.csv(res, file ="deseq_results.csv")
```

## 2. Pathway Analysis

Here we are going to use the `gage` package for pathway analysis. Once we have a list of enriched pathways, we're going to use the `pathview` package to draw pathway diagrams, shading the molecules in the pathway by their degree of up/down-regulation.

### KEGG pathways

```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```
The main **gage()** function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

Note that we used the **mapIDs()** function above to obtain Entrez gene IDs (stored in res$entrez) and we have the fold change results from DESeq2 analysis (stored in res$log2FoldChange).


```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```
Now, let's run the **gage** pathway analysis

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
```
```{r}
head(keggres$less)
```
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```
```{r}
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```
```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

> Q7. Repeat but with top 5 down-regulated genes

```{r}
keggresdown <-substr(rownames(keggres$less)[1:5], start=1, stop=8)
keggresdown
```
```{r}
#pathview(gene.data=foldchanges, pathway.id=keggresdown, species="hsa")
```

The tables saved as a png in my directory, but I have always been unable to upload to quarto document smoothly.

## 3. Gene Ontology

We can do a similar procedure with Gene Ontology

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

## 4. Reactome Analysis

Reactome is database consisting of biological molecules and their relation to pathways and processes. 

Let's now conduct over-representation enrichment analysis and pathway-topology analysis with Reactome using the previous list of significant genes generated from our differential expression results above.

First, Using R, output the list of significant genes at the 0.05 level as a plain text file:

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

> Q8. What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The most significant (lowest) p-value is for the cell cycle pathway. This does actually match pretty well with the KEGG results, which is impressive considering they are entirely different methods. The differences between the two methods still exist though, which is likely due to them receiving their data from (somewhat) different sources.





