---
title: "Class 7: Machine Learning 1"
author: "Peter"
format: html
toc: true
---

Today we will explore unsupervised machine learning methods starting with clustering and dimensionality reduction.

## Clustering

To start let's make up some data to cluster where we know what the answer should be. The `rnorm()` function will help us here.

```{r}
hist(rnorm(10000, mean=3))
```
Return 30 numbers centered on -3 and make a plot

```{r}
tmp <- c(rnorm(30, mean=-3), rnorm(30, mean=3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```
### K-means

The main function in base R for k-means clustering is called `kmeans()`:

```{r}
km <- kmeans(x, centers=2)
km
```
The `kmeans()` function returns a "list" of 9 components. You can see the named components of any list with the `attributes` function.

```{r}
attributes(km)
```
> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. Cluster assignment vector?

```{r}
km$cluster
```

> Q. Cluster centers?

```{r}
km$centers
```

> Q. Make a plot of our `kmeans` results sharing cluster assignment using different colors for each cluster/group of points and cluster centers in blue.

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15, cex=2)
```
> Q. Run `kmeans()` again and this time cluster into 4 groups and plot the same result figure as above

```{r}
km4 <- kmeans(x, centers=4)
plot(x, col=km4$cluster)
points(km4$centers, col="blue", pch=15, cex=2)
```
> **key-point**: K-means clustering is super popular but can be misused. One big limitation is that it can impose a clustering pattern on your data even if clear natural grouping doesn't exist - i.e. it does what you tell it to do it terms of `centers`.

### Hierarchical Clustering

The main function in base R for hierarchical clustering is called `hclust()`.

You can't just pass our dataset as is into `hclust()`; you must give "distance matrix" as input. We can get this from the `dist()` function in R

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```
The results of `hclust()` don't have a useful `print()` method but do have a special `plot()` method.

```{r}
plot(hc)
abline(h=8, col="red")
```
To get our main cluster assignment (membership vector) we need to cut the tree at the goal posts... 

```{r}
grps <- cutree(hc, h=8)
grps

table(grps)
```
Hierarchical Clustering is distinct in that the dendrogram can reveal the potential grouping in your data (unlike K-means) 😩

## Principal Component Analysis (PCA)

PCA is a common and highly useful dimensionality reduction technique used in many fields - particularly bioinformatics.

Here we will analyse some data from the UK on food consumption

### Data import

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)
```

```{r}
head(x)
```
```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

We can also use a pairs plot

```{r}
pairs(x, col=rainbow(nrow(x)), pch = 16)
```

### PCA to the rescue

The main function in base R for PCA is called `prcomp()`.

```{r}
pca <- prcomp(t(x))
summary(pca)
```
The `prcomp()` function returns a list object of our results with five attributes

```{r}
attributes(pca)
```
The two main results in here are `pca$x` and `pca$rotation`. The first of these (`pca$x`) contains the scores of the data on the new PC axis - we use these to make our "PCA plot".

```{r}
pca$x
```
```{r}
library(ggplot2)
library(ggrepel)
ggplot(pca$x) + 
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point() +
  geom_text_repel()
```

The second major result is contained in the `pca$rotation` object. Let's plot this to see what PCA is picking up...

```{r}
ggplot(pca$rotation) + 
  aes(PC1, rownames(pca$rotation)) + 
  geom_col()
```

The above plot shows how N.Ireland fares compared to Great Britain
